Unified visualization of seasonal influenza evolution in the laboratory and in nature.

### 2019-06-11 paper/paper.tex
add figures to paper

\section{Related Work}

The Bloom lab at Fred Hutch performs are large amount of DMS work to investigate functional stability and immune escape and selection for important human pathogens such as HIV, influenza, and Zika virus. Originally developed by Thyagarajan and Bloom \cite{thyagarajan2014inherent}, initial DMS experiments on influenza virus hemagglutinin focused on understanding which sites of the protein could tolerate multiple amino acids and cause a functional infection. These studies have been extended to include not only purifying selective forces, but also immune selection as well, by innate immunity \cite{ashenberg2017deep} and adaptive immunity involving antibody response \cite{Lee2018}. In total, these experiments have helped us to define broad atlases of paths for escaping immunity \cite{dingens2019antigenic}. Importantly, the development of such atlases, and the visualization of these data, requires an intensive computational workflow that involves iteratively generating static graphs for the relevant streams of data: immune selection as measured by the DMS experiment and the 3D structure of the protein. By default these images also do not link a third relevant data stream, and that is the frequencies of amino acids observed at different sites of the protein in nature. Thus the task of our visualization platform is to provide an easy way to view and interact with these three data streams in a linked fashion that is also easy to use in the absence of coding expertise.

\section{Methods}

This visualization platform links three data views together by their common shared element, site in the protein. The three data streams are: 1) DMS differential selection value, which is a quantitative measure of the preference for a certain amino acid at a specific site in the protein, 2) the natural amino acid frequencies of a site, a quantitative variable the describes what proportion of all sampled strains of influenza at time T have a specific amino acid at a site, and 3) the 3-D structure of protein itself, which takes linear sequence space and folds that up to reflect where certain sites of the sequence fall on the protein structure. We describe design decisions about how to visualize these data streams below. Our visualization was implemented using D3 \cite{bostock2011d3}

\begin{figure}[H]
	\includegraphics[width=1.0\textwidth]{viz-overview.png}
	\caption{\textbf{Full view of visualization platform}
   The initial view of the visualization shows three panels: the laboratory-based DMS data, influenza variant frequencies in nature, and the 3D protein structure.
	}
	\label{clade-2-detail}
\end{figure}

@@

@@

@@

>>>>

\section{Related Work}
The Bloom lab at Fred Hutch performs are large amount of DMS work to investigate functional stability and immune escape and selection for important human pathogens such as HIV, influenza, and Zika virus. Originally developed by Thyagarajan and Bloom \cite{thyagarajan2014inherent}, initial DMS experiments on influenza virus hemagglutinin focused on understanding which sites of the protein could tolerate multiple amino acids and cause a functional infection. These studies have been extended to include not only purifying selective forces, but also immune selection as well, by innate immunity \cite{ashenberg2017deep} and adaptive immunity involving antibody response \cite{Lee2018}. In total, these experiments have helped us to define broad atlases of paths for escaping immunity \cite{dingens2019antigenic}. Importantly, the development of such atlases, and the visualization of these data, requires an intensive computational workflow that involves iteratively generating static graphs for the relevant streams of data: immune selection as measured by the DMS experiment and the 3D structure of the protein. By default these images also do not link a third relevant data stream, and that is the frequencies of amino acids observed at different sites of the protein in nature. Thus the task of our visualization platform is to provide an easy way to view and interact with these three data streams in a linked fashion that is also easy to use in the absence of coding expertise.

\section{Methods}
This visualization platform links three data views together by their common shared element, site in the protein. The three data streams are: 1) DMS differential selection value, which is a quantitative measure of the preference for a certain amino acid at a specific site in the protein, 2) the natural amino acid frequencies of a site, a quantitative variable the describes what proportion of all sampled strains of influenza at time T have a specific amino acid at a site, and 3) the 3-D structure of protein itself, which takes linear sequence space and folds that up to reflect where certain sites of the sequence fall on the protein structure. We describe design decisions about how to visualize these data streams below. Our visualization was implemented using D3 \cite{bostock2011d3}

\begin{figure}[H]
	\includegraphics[width=1.0\textwidth]{viz-overview.png}
	\caption{\textbf{Full view of visualization platform.}
   The initial view of the visualization shows three panels: the laboratory-based DMS data, influenza variant frequencies in nature, and the 3D protein structure.
	}
	\label{overview}
\end{figure}

@@

\begin{figure}[H]
	\includegraphics[width=1.0\textwidth]{dms-brushed-tooltip.png}
	\caption{\textbf{Visualization of differential selection data generated from DMS experiments.}
  The laboratory data can be explored and zoomed in on by brushing the context panel below the primary panel, giving both detail and overview to the user. Sites that are colored black in this panel represent sites for which there is natural frequency data. For all grey sites, only one variant of influenza at that site is found to circulate naturally. Mousing over a tip brings up details on demand, and clicking on a data point selects that site for visualization in the other panels, and colors the point according to it's maximal differential selection value.
	}
	\label{dms-data}
\end{figure}

@@

\begin{figure}[H]
	\includegraphics[width=1.0\textwidth]{frequencies-panel.png}
	\caption{\textbf{Frequencies of influenza variants in nature.}
  Once a site has been selected from the laboratory data panel, this panel shows the frequencies of influenza viruses with different amino acids at that site over the last three years. In this example, at site 193 we see that there are two variants circulating, one has a phenylalanine (F) at this site, while the other has a serine (S). We can see that the phenyalanine variant is at higher frequencies, but that the serine variant is rising in frequency.
  }
	\label{frequencies-panel}
\end{figure}

@@

\begin{figure}[H]
	\includegraphics[width=1.0\textwidth]{3d-structure-panel.png}
	\caption{\textbf{3D protein structure with selected site highlighted.}
  3D protein structure panel, with the selected site colored according to it's differential selection value. The tooltip also provides information about which amino acid is at highest frequency at the site. In this image, site 193 has a PHE, or phenylalanine.
  }
	\label{3d-structure}
\end{figure}
